Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-05-13 ltx/data_paper/DataPaper.tex
edits to EM section

	\boldsymbol\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0}, \\
\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1}, \\
\boldsymbol\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).

@@

 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)

@@

 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\right)

@@

substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the estimated state $\hat{\mathbf{x}}_t$ 

@@

 q_L(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right) d\mathbf{r}'

@@

Define
	e(x) = q(x) - q_L(x) \nonumber

@@

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, by substituting  \ref{eq:qTaylor} into equations \ref{eq:Xi3}-\ref{eq:Xi5} an approximate to the non-linear $\boldsymbol\Xi$-variables can be calculated which provides an approximate lower bound on the log-likelihood function (see supporting information), 
\begin{align}
\boldsymbol\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t\right] \\
 \boldsymbol\Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \end{align}
where
\begin{align}\label{eq:lambdat} 
	{\boldsymbol \lambda}_t^{(i)}&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'
\end{align}
and
\begin{equation}
 \boldsymbol\Lambda_t=\sum_{i,j=1}^{n_x}[\mathbf{P_t}]_{ij}{\boldsymbol\lambda}_t^{(i)\top}\tilde{\boldsymbol\Sigma}_e^{-1}{\boldsymbol\lambda}_t^{(j)}
\end{equation}

Given the state estimates, $\mathbf {\hat x}_t$ the state covariance $\mathbf P_t$ and cross-covariance $\mathbf M_t$ , the $\boldsymbol\Xi$-variables can be computed directly. In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases. The computational complexity of the algorithm is given in Table~\ref{table:ComputationalComplexity}. In this table only the computational complexity of variables specific to the presented algorithm are shown, and the computational complexity of the URTSS can be found in \cite{Sarkka2010}.

@@

The objective of the M-step is to maximize a lower bound on the joint likelihood function, $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$, where $\mathbf X$ is the set of all states, $\mathbf Y$ is the set of all observations, and $\boldsymbol \Theta$ denotes the set of all unknown parameters, i.e., connectivity kernel parameters $\boldsymbol\theta$, synaptic dynamic $\xi$, disturbance variance $\sigma^2_d$ and the observation noise variance $\sigma^2_{\epsilon}$.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelBasis1.pdf}
\end{center}
\caption{{\bf Connectivity kernel basis functions, $n_{\theta}=25$}, $\sigma^2_{\theta}=3.5$.}
\label{fig:KernelWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/DisturbanceWidthEstimation.pdf}
\end{center}
\caption{{\bf Disturbance support estimation}.}
\label{fig:DisturbanceWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimate.pdf}
\end{center}
\caption{{\bf Connectivity kernel estimates over 100 realisations.} left: True kernel; right;Estimated kernel.}
\label{fig:KernelEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimateCrossSection.pdf}
\end{center}
\caption{{\bf Cross-section of connectivity kernel estimates over 100 realisation.} True is blue; Estimate is green; 95\% Confedence interval is red.}
\label{fig:KernelEstimateCrossSection}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Histogram.pdf}
\end{center}
\caption{{\bf Synaptic dynamic parameter and hyperparameter estimates}.}
\label{fig:Histogram}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=1]{./Figures/FieldEstimates.pdf}
\end{center}
\caption{{\bf Field Estimates}. }
\label{fig:FieldEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Q.pdf}
\end{center}
\caption{{\bf Lower-bound of the log-likelihood}. $Q(\boldsymbol\Theta',\boldsymbol\Theta')=-(T-1)\left[n_y(1+\ln \hat{\sigma}^2_{\epsilon})+ n_x(1+\ln \hat{\sigma}^2_d)\right]$.}
\label{fig:LowerBound}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

>>>>

	\boldsymbol\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}^b+\mathbf{\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\right)\label{eq:defofXi0}, \\
\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right) \label{eq:defofXi1}, \\
\boldsymbol\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_t^{b\top}\right).

@@

 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t^b\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t^b\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)

@@

 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right)\right)

@@

substituting \eqref{eq:SigmoidTaylor} in \eqref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the smoothed state $\hat{\mathbf{x}}_t^b$ 

@@

 q_L(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t^b\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t^b\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right) d\mathbf{r}'

@@

Having approximated $q\left(\mathbf x_t\right)$, by substituting  \eqref{eq:qTaylor} into equations (\ref{eq:Xi3}-\ref{eq:Xi5}) an approximate to the non-linear $\boldsymbol\Xi$-variables can be calculated which in turn provides an approximate lower bound on the log-likelihood function (see supporting information), 
\begin{align}
\boldsymbol\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\Lambda_t\right] \\
 \boldsymbol\Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \end{align}
where
\begin{align}\label{eq:lambdat} 
	{\boldsymbol \lambda}_t^{(i)}&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t)d\mathbf r'
\end{align}
and
\begin{equation}
 \boldsymbol\Lambda_t=\sum_{i,j=1}^{n_x}[\mathbf{P_t}]_{ij}{\boldsymbol\lambda}_t^{(i)\top}\tilde{\boldsymbol\Sigma}_e^{-1}{\boldsymbol\lambda}_t^{(j)}
\end{equation}

Given the state estimates, $\mathbf {\hat x}_t$ the state covariance $\mathbf P_t$ and cross-covariance $\mathbf M_t$ , the $\boldsymbol\Xi$-variables can be computed directly.  The computational complexity of the algorithm is given in Table~\ref{table:ComputationalComplexity}. In this table only the computational complexity of variables specific to the presented algorithm are shown, and the computational complexity of the URTSS can be found in \cite{Sarkka2010}.

Define
\begin{equation}
	e(x) = q(x) - q_L(x) \nonumber
\end{equation}

@@

@@

% In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.
The objective of the M-step is to maximize the lower bound found in E-step, whcich consequenly increases the log-likelihood function $p(\mathbf Y;\boldsymbol\Theta)$.
% on the joint likelihood function, $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$, where $\mathbf X$ is the set of all states, $\mathbf Y$ is the set of all observations, and $\boldsymbol \Theta$ denotes the set of all unknown parameters, i.e., connectivity kernel parameters $\boldsymbol\theta$, synaptic dynamic $\xi$, disturbance variance $\sigma^2_d$ and the observation noise variance $\sigma^2_{\epsilon}$.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelBasis1.pdf}
\end{center}
\caption{{\bf Connectivity kernel support estimate.} The estimated support of the kernel is shown by the red circle. Black area shows the non-zero values of the support after thresholding. A $5\times5$ grid of basis functions (blue circles) used to reconstruct the connectivity kernel.}
\label{fig:KernelWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/DisturbanceWidthEstimation.pdf}
\end{center}
\caption{{\bf Disturbance support estimation}.}
\label{fig:DisturbanceWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimate.pdf}
\end{center}
\caption{{\bf Connectivity kernel estimates over 100 realisations.} left: True kernel; right;Estimated kernel.}
\label{fig:KernelEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimateCrossSection.pdf}
\end{center}
\caption{{\bf Cross-section of connectivity kernel estimates over 100 realisation.} True is blue; Estimate is green; 95\% Confedence interval is red.}
\label{fig:KernelEstimateCrossSection}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Histogram.pdf}
\end{center}
\caption{{\bf Synaptic dynamic parameter and hyperparameter estimates}.}
\label{fig:Histogram}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=1]{./Figures/FieldEstimates.pdf}
\end{center}
\caption{{\bf Field Estimates}. }
\label{fig:FieldEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Q.pdf}
\end{center}
\caption{{\bf Lower-bound of the log-likelihood}. $Q(\boldsymbol\Theta',\boldsymbol\Theta')=-(T-1)\left[n_y(1+\ln \hat{\sigma}^2_{\epsilon})+ n_x(1+\ln \hat{\sigma}^2_d)\right]$.}
\label{fig:LowerBound}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
